---
title: "ENVS 193DS Final"
author: "Angela Larson"
date: "8 June 2025"
format:
  html:
    toc: true # use this to display a table of contents
execute:
  message: false # use this to make sure messages don't show up
  warning: false # use this to make sure warnings don't show up
---

[Repo Link]: https://github.com/angelalarson/ENVS-193DS_spring-2025_final.git

```{r to read in packages and data}
#| message: false
library(tidyverse) # general use
library(here) # file organization
library(janitor) # cleaning data frames
library(cowplot) # for plot theme
library(scales) # modifying axis labels
library(ggeffects) # getting model predictions
library(MuMIn) # model selection
library(DHARMa)
library(ggplot2)


sst <- read_csv(here("data", "SST_update2023.csv")) # reading in SB channel SST data
nest_boxes <- read_csv(here("data", "occdist.csv")) # reading in nest box data
```

# Problem 1 - Research writing

## a. Transparent statistical methods

In part one, they used a pearson correlation test. In part two, they used a one-way ANOVA.

## b. More information needed

Since the ANOVA returned at least two or more of the variables are statistically different from each other, we need to run a Tukey's post-hoc test to determine which specific nitrogen-containing variables are statistically different from each other. The ANOVA told us a statistically significant difference exists within the variables (urban land, atmospheric deposition, fertilizer, wastewater treatment, and grasslands), the post-hoc identifies between which variables.

Furthermore, we would run an eta squared test (n^2^) to determine the effect size. The effect size determines how much the grouping variable (urban land, atmospheric deposition, fertilizer, wastewater treatment, and grasslands) explains the response variable (average nitrogen load), given by the size of the variation in the response variable as described by the grouping variable. To put it simply, it describes how big of a difference exists between the variables, not just the significance. 

## c. Suggestions for re-writing

We determined there was a correlation between distance from headwater (km) and annual total nitrogen load (kg year{-1}) (Pearson correlation test, r = **correlation coefficient**, p = 0.03, $\alpha$ = **significance level**). 
We found a small/medium/large ($\eta^2$ = **n^2^**) difference between sources (urban land, atmospheric deposition, fertilizer, wastewater treatment, and grasslands) in average nitrogen load (kg year{-1}), (one-way ANOVA, F(**among groups df, within groups df**) = **F-statistic**, p = 0.02, $\alpha$ = **significance level**). On average, **source** tended to have smaller/larger nitrogen loads than **source** (Tukey's HSD: p < **p-value**).

# Problem 2 - Data visualization

## a. Cleaning and summarizing
```{r cleaning and summarizing}
#| message: false

sst_clean <- sst %>% # creating new data frame called sst_clean from sst
  select(temp, date) %>%  # selecting only temp and date columns
  filter(date > as_date("2017-12-31")) %>% # only including dates of 2018-2023
  
 mutate(month = factor(month(date, label = TRUE, abbr = TRUE), # extract month from each date, labels it, and abbreviates the label
                       levels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", 
                              "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")), # makes the months a categorical factor
        
        year = factor(as.character(year(date)), #set year as factor and extract year from the date column
                       levels = c("2018", "2019", "2020", "2021", "2022", "2023"))) %>%
  
            group_by(year, month) %>%  # group by year and month
  summarize(mean_monthly_sst = round(mean(temp), 1)) %>% # calculate the mean sst per month
  ungroup() # add to ungroup grouped variables and match structure provided in example

slice_sample(sst_clean, n = 5) # displaying five rows of clean data
str(sst_clean) # showing the structure
```

## b. Visualize the data
```{r visualize the data}
#| message: false

ggplot (data = sst_clean, # using sst_clean data frame
        aes(x = month, # assigning x axis to be month
            y = mean_monthly_sst, # assigning y axis to be mean monthly sst
            color = year)) + # assigning color to be organized by year
  
  geom_point() + # adding points
  geom_line(aes(group = year, color = year)) + # adding a line and color for each year
  
scale_color_manual(values = c("#c6dbef", # manually adjusting the colors
                              "lightskyblue3", 
                              "#6baed6", 
                              "#4292c6", 
                              "#2171b5", 
                              "#084594")) + 
  
  scale_y_continuous(breaks = c(14, 16, 18, 20)) + # supposed to add the 20th tick mark but is not

  labs(x = "Month", # name the x axis
       y = "Mean monthly sea surface temperature (°C)", # name the y axis
       color = "Year") + # name the legend
  theme_cowplot() + # using cowplot theme
  theme(axis.line = element_line(color = "slategray4", linewidth = 0.2), # changing color and width of axes
        axis.ticks = element_line(color = "slategray4"), # changing color of axis tick marks
        axis.text = element_text(color = "slategray4"), # changing the color of axis text
        legend.position = c(0.05, 0.75), # changing legend position
        panel.border = element_rect(color = "slategray4", linewidth = 0.7)) # adding a border
  
```
# Problem 3 - Data analysis
```{r cleaning data}
nest_boxes_clean <- nest_boxes %>% 
  clean_names() %>% 
  mutate(season = factor(season))
```
## a. Response variable

The response variable is Swift Parrot box occupancy (binary variable "sp"), where 1 indicates a box occupied by a Swift Parrot. 0 would represent occupancy by the Common Starling or Tree Martin, or not occupied at all. 

## b. Purpose of study

The Swift Parrot, a critically endangered species and the focus of restoration efforts, is the target species of this restoration-based study that aims to determine whether nest boxes promote nesting of the target species or of its competitors, the Common Starling and Tree Martins. If competitors nest in the boxes, it could potentially impact the Swift Parrot's nesting success and project goals. 

## c. Difference in “seasons”

This study compares two seasons: 2016, when the nest boxes were first installed, and 2019, three years later. This time gap allows the author to examine changes in nest box use over time, specifically whether or not competitor species occupy more boxes meant for the Swift Parrot over the three years. 

## d. Table of models

| Model number | Season |  Forest Edge Distance    | Model Description            | 
|:------------:|:------:|:------------------------:|------------------------------|  
| 0            |        |                          | no predictors (null model)   |
| 1            |    X   |             X            | both predictors              | 
| 2            |    X   |                          | season only                  |   
| 3            |        |             X            | forest edge distance    only |    

## e. Run the models

```{r, model-fitting}
#| message: false
#| echo: false

results = 'hide'
# model 0: null model
model0 <- glm(
  sp ~ 1, # no predictors
  data = nest_boxes_clean, 
  family = "binomial" # include for binomial data
)

# model 1: all predictors
model1 <- glm(
  sp ~ season + edge_distance, # full model 
  data = nest_boxes_clean,
  family = "binomial" # include for binomial data
)

# model 2: season only
model2 <- glm(
  sp ~ season, # season only
  data = nest_boxes_clean,
  family = "binomial" # include for binomial data
)

# model 3: forest edge distance only
model3 <- glm(
  sp ~ edge_distance, # only edge distance
  data = nest_boxes_clean,
  family = "binomial" # include for binomial data
)
```

## f. Check the diagnostics

```{r, model-diagnostics}
#| message: false

par(mfrow = c(2,2)) # creates a 2x2 grid to display diagnostic plots THIS IS NOT WORKING
# Model 0
plot(
  sim0 <- simulateResiduals(model0)
)
# Model 1
plot(
  sim1 <- simulateResiduals(model1) # redisuals look ok for best model
)
# Model 2
plot(
  sim2 <- simulateResiduals(model2)
)
# Model 3
plot(
  sim3 <- simulateResiduals(model3)
)

```

## g. Select the best model

```{r, model-selection}
AICc(model0,
     model1, # lowest AIC, best model
     model2,
     model3) %>% 

  arrange(AICc) # arranging output in descending order of AIC
```

The best model, based on Akaike’s Information Criterion (AIC), included season, forest edge distance, and their interaction as predictors of Swift Parrot nest box occupancy. The model including both season and forest edge distance had the lowest AIC value, indicating the best trade-off between model simplicity and fit among the models tested.

## h. Visualize the model predictions

```{r, model-predictions}
#| message: false

model1_predictions <- ggpredict( model1, # naming model object
  terms = c("edge_distance", "season") # listing predictors
) %>% 
  rename(
    edge_distance = x,
    season = group
  )

view(model1_predictions)
```


```{r, model-visualization}

ggplot(data = nest_boxes_clean, # using nest_boxes_clean data frame
       aes(x = edge_distance, # x is edge-distance
           y = sp, # y axis
           color = season)) + # color grouped by season
  
  geom_point(size = 3, #change size
            alpha = 0.4, # adjust transparency
             color = "slateblue4") + # change color
  
  # Confidence ribbon with 95% CI
 geom_ribbon(data = model1_predictions, # using model1_predictions
             aes(x = edge_distance, # x is edge-distance
                  y = predicted, # y is predicted occupancy probability
                  ymin = conf.low, # lower CI bound
                  ymax = conf.high, # upper CI bound
                  fill = season), # full ribbons by season
              alpha = 0.4, # adjusting transparency
              color = NA) + # remove border on ribbon
  

  geom_line(data = model1_predictions, # use model prediction data
            aes(x = edge_distance, # x is edge-distance
                y = predicted, # y is predicted occupancy probability
                color = season), # group color by season
            size = 1.2) + # change line thickness
  
 scale_color_manual(values = c("2016" = "violetred", "2019" = "gold2")) + # changing colors to associate with season
  scale_fill_manual(values = c("2016" = "violetred", "2019" = "gold2")) +

  labs(
    x = "Distance to forest edge (m)", #  axis label
    y = "Probability of Swift Parrot occupancy", # y axis label
    color = "Season", # legend title for line color
    fill = "Season", # legend title for ribbon color 
    title = "Swift Parrot nest box occupancy as determined by distance to forest edge and year" # title label
  ) +
  theme_cowplot() + # use cowplot theme to remove lines and such
  theme(legend.position = c(0.8, 0.8), # place legend inside x axis
         plot.title = element_text(size = 11)  # change title font size
) 
  
```

## i. Write a caption for your figure.

**Figure 1. Swift Parrot Nest Box Occupancy Decreases with Distance from Forest Edge and Differs between Seasons**
Data from Stojanovic et al. (2020).
Points represent the observations of Swift Parrot (*Lathamus discolor*) nest box occupancy (0 = unoccupied, 1 = occupied) and distance to forest edge (m) in different seasons (total n = 227).
Lines show the model-predicted occupancy probabilities based on a binomial GLM with 95% confidence intervals.
Colors represent season (orange: 2016, blue: 2019).

## j. Calculate model predictions

```{r, model1-predictions-0-and-900}
# Generating model prediction for specific distances for each season
model1_prediction_0_and_900 <- ggpredict(
  model1,
  terms = c("edge_distance [0,900]", "season")
)

# View the output
model1_prediction_0_and_900
```

## k. Interpret your results

At the edge of the forest (0 m), the predicted probability of Swift Parrot nest box occupancy was **higher in 2016 than in 2019**, suggesting that the occupancy of the nest box was initially strong. However, at 900 m from the edge of the forest, the occupancy probabilities were **low in both seasons**, with a slightly larger model uncertainty for 2019; overall, this indicates that the distance from the edge of the forest **negatively influences** Swift Parrot occupancy.
The model and figure display a **decreasing trend** in Swift Parrot occupancy with **increasing the distance from the edge of the forest**, which is consistent across both seasons, although the decline appears to be steeper in 2016.
Biologically, the trends likely reflect the Swift Parrot's **natural nesting preferences**, which is that they're edge-adapted cavity nesters, so their probability of occupying a nest box is **highest at the edge of the forest**. In the paper, we see that Common Starlings increasingly dominated the nest boxes closest to the forest edge over time, indicating that they outcompeted the other birds for these prime sites. Over time, we see that Tree Martins have learned to exploit interior boxes, further reducing the availability of suitable nesting sites for parrots in more more interior locations.


# Problem 4 - Affective and exploratory visualizations

## a. Comparing visualizations

I decided to represent the data in Homework 2 as a line graph and box plots, and didn't really get any clear visuals of the data at the time. By grouping the data into four different time slots and representing them as bar graphs, I was able to see visual patterns appear between step count and time on campus.

There aren't many similarities, however I similarly saw that grouping steps by exercise type really showed no correlation between average step count and exercise type. 

The most clear pattern is in the affective visualization as the average number of steps is somewhat correlated to time on campus; however, all the standard error bars overlap with each other so that correlation is only observed and would not be supported statistically. I plotted number of steps against time on campus as a line graph in homework 2, which showed a similar correlation as to what was observed in the final visualization. However, with so few points in homework 2, it is hard to compare as the data was quite variable at the time.

I received primarily positive feedback, however it was suggested that I incorporate color into my drawing to visually distinguish between mean and error bars and buildings. I incorporated color to make it as visually distinguishable as possible.

## b. Sharing your affective visualization

This was done during section on Friday.
